Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-10-24 _tex/manuscript.tex
Built site for gh-pages

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

When treating head and neck squamous cell carcinomas (HNSCC) with
radiotherapy or surgery, not only the primary tumor and the clinically
detected lymph node metastases are targeted. Per current guidelines,
large volumes of the neck are part of the the elective clinical target
volume (CTV-N)
\citep{gregoire_ct-based_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensity-modulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}.
This is to minimize the risk of regional recurrences due to untreated
microscopic disease which cannot be detected by current in-vivo imaging
modalities such as computed tomography (CT), magnetic resonance imaging
(MRI), or positron emission tomography (PET). However, minimizing the
risk of missing occult disease in the lymph drainage region must be
balanced against the toxicity related to unnecessary treatment of
healthy tissue.

The aforementioned guidelines for the definition of the CTV-N are based
on anatomically defined lymph node levels (LNL)
\citep{gregoire_delineation_2014} and the overall prevalence of lymph
node metastases in the LNLs. Based on this, they recommend extensive
irradiation of both sides of the neck for a majority of patients.
However, this prevalence the guidelines are based on does not take into
account an individual patient's state of tumor progression, which may
have a great impact on a patient's personal risk for occult disease in a
given LNL. For example, a patient with substantial nodal involvement in
the ipsilateral side and an advanced primary tumor that crosses the
mid-sagittal plane would receive extensive elective irradiation of the
contralateral LNLs II, III, and IVa. However, if the patient presented
with no clinically detectable nodal disease (cN0) and only a small,
clearly lateralized T1 tumor, the same contralateral CTV-N would be
recommended \citep{biau_selection_2019}.

To better quantify this personalized risk of occult disease we have
previously developed an intuitive, probabilistic hidden Markov model
\citep{ludwig_hidden_2021, ludwig_modelling_2023} which was originally
based on a conceptually similar Bayesian network model
\citep{pouymayou_bayesian_2019}. So far, these models were limited to
the ipsilateral side of the neck. The main contribution of this work
will be to extend the formalism to include risk predictions for the
contralateral side as well. We believe this holds the potential to
reduce the electively irradiated volumes in the contralateral neck for
patients whose individual risk for occult disease is low. In turn, this
may substantially reduce the toxicity from irradiation and the
associated decrease in quality of life.

@@

To be able to create models for lymphatic tumor progression that include
all relevant LNLs including the contralateral side, we have collected a
detailed dataset of 833 patients with newly diagnosed oropharyngeal
squamous cell carcinomas
\citep{ludwig_dataset_2022, ludwig_multi-centric_2023}. It reports the
lymph node involvement per LNL of every patient individually in tabular
form, in addition to other primary tumor and patient characteristics
such as T-category, subside, and lateralization of the primary tumor,
and HPV p16 status. Their patient records have been collected at four
different institutions and a brief overview over some of their patients'
characteristics are shown in table~\ref{tbl-data-overview}. The data
from the Inselspital Bern and the Centre Léon Bérard only consist of
patients who received a neck dissection. Since this treatment is more
commonly chosen for early T-category patients, they also make up a
larger portion of the respective dataset. The majority of patients from
the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital
was treated with definitive radiotherapy. For a subset of 83 patients in
the CLB dataset, the lateralization of the primary tumor was not
reported.

@@

\caption{}\label{T_8749e}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

Pathological involvement is thus available only for a subset of the
patients and only for the dissected levels. For the remaining patients,
only clinical involvement is available. For the analysis performed in
this paper, all available diagnostic information has been combined into
a consensus decision representing the most likely state of involvement
for each patient and LNL, using literature values for sensitivity and
specificity of different modalities
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.
These values are tabulated in table~\ref{tbl-spec-sens}.

This consensus was formed as follows: Suppose the involvement status
\(X_v\) of LNL \(v\) was assessed using different diagnostic modalities
\(\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}\),
each characterized by their own pair of sensitivity and specificity
values \(s_N^{\mathcal{o}}\) and \(s_P^{\mathcal{o}}\), with
\(\mathcal{o} \in \mathcal{O}\). Then we have \(|\mathcal{O}|\)
observations \(z_v^{\mathcal{o}} \in \left[ 0, 1 \right]\), where 0
stands for ``healthy'' and 1 for ``involved''. We can then compute the
most likely true involvement \(X_v\) using the likelihood function

@@

The entire data, including additional patients with tumors in other
primary locations than the oropharynx, is publicly available: It may be
\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
it can be interactively explored too,
\href{https://github.com/rmnldwg/lydata}{from GitHub},
\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
\citet{ludwig_multi-centric_2023}. The Data-in-Brief publications
contain a detailed description of the datasets and the data format.
However, these publications do not include the most recent dataset
addition from Vall d'Hebron Barcelona Hospital.

@@

These datasets allow us to investigate correlations between the
involvement of individual LNLs, or between risk factors and involvement.
In figure~\ref{fig-data-strat}, we have plotted the prevalence of each
contralateral LNL's involvement, stratified by risk factors such as
T-category, number of ipsilaterally involved LNLs, and whether the tumor
extended over the mid-sagittal line. We also show in the fourth subfigre
that T-category and ipsilateral involvement are not just rpedictive of
contralateral involvement because they correlate with midline extension.
A similar but more complete stratification is also tabulated in the
appendix in table~\ref{tbl-data-strat}.

@@

\includegraphics{manuscript_files/mediabag/figures/fig_data_strat.pdf}

@@

The top left panel in figure~\ref{fig-data-strat} indicates that
T-category is correlated with contralateral involvement (as it is with
overal involvement). This is simply because T-category may on average be
considered a surrogate for the time between onset of disease and
diagnosis. I.e., a patient with a T4 tumor was -- on average --
diagnosed later than a patient with a T1 tumor. Thus, the former did
have more time to develop metastases.

Similarly, ipsilateral involvement correlates with contralateral
metastasis. The tumor of a patients with many metastases in ipsilateral
LNLs was probably able to spread for longer (or faster) compared to a
tumor in a patient with no nodal disease. This, too, may therefore be
considered a surrogate for the duration of the disease. In addition, it
has been hypothesized that bulky nodal disease ipsilaterally may also
redirect lymph fluids to the contralateral side.

The bottom left panel in figure~\ref{fig-data-strat} shows that patients
with a tumor crossing the mid-sagittal line show contralateral
involvement vastly more often compared to patients with clearly
lateralized tumors. This makes intuitive sense, because the lymphatic
system in the head and neck region is typically symmetric and thus no
major vessels cross the midline. Therefore, interstitial fluids from the
primary tumor -- which we assume to carry living malignant cells -- may
only reach the blind-ended lymphatic vessels in the contralateral neck
via short-ranged diffusion. Which in turn is only possible when the
primary tumor is close enough to the mid-sagittal line or crosses it.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  More advanced T-category should lead to higher risk for nodal disease.
  One approach to achieve this via the expected time of diagnosis has
  already been developed in the form of a hidden Markov model
  \citep{ludwig_hidden_2021}.
\item
  The degree of ipsilateral involvement should give the model
  information on the time that may have passed between onset and
  diagnosis of the disease. This should come in addition to what can be
  inferred about the time from T-category alone.
\item
  A tumor that extends over the mid-sagittal line should yield
  contralateral metastases with much higher probability.
\end{enumerate}

Over the course of this work, we will first briefly recap the HMM in
section~\ref{sec-unilateral}, which was so far used to model ipsilateral
lymphatic progression only. Then, we intuitively extend it to include
the contralateral side as well in section~\ref{sec-ext-to-contra}. In
this section, we also introduce a way of modelling the tumor's midline
extension as a random variable (section~\ref{sec-midline}) and lastly
talk about how it may affect the contralateral spread in
section~\ref{sec-params-symmetry}.

@@

\includegraphics[width=0.4\textwidth,height=\textheight]{manuscript_files/mediabag/static/full-graph.pdf}

@@

Using \{\#eq-bilateral-marginal\} along with
equation~\ref{eq-lambda-matrix}, we can write the above distribution
algebraically as a product:

@@

\begin{equation}\phantomsection\label{eq-uni-bayes-law}{
P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid d, \boldsymbol{\hat{\theta}} \big)
= \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
}\end{equation}

@@

\subsection{Involvement Data
Consensus}\label{involvement-data-consensus}

@@

\includegraphics{manuscript_files/mediabag/figures/fig_model_burnin_history.pdf}

@@

In table~\ref{tbl-midline-params}, we tabulate the mean and standard
deviation of the sampled parameters for the full midline model.
Considering the parameters of the bilateral model that describe the
ipsilateral spread (\(b^i_v\), \(t_{vr}\) and \(p_\text{adv.}\)), we
note that the bilateral model mostly reproduces the parameter values
reported in the earlier publication \citep{ludwig_modelling_2023} for
the unilateral model. The small differences may be mostly due to the
differences in training dataset. Therefore, the analysis of the models
capability to adequately describe ipsilateral lymph node involvement is
not repeated here and we focus instead on the analysis of contralateral
involvement. We can further note that the contralateral spread
parameters \(b^c_v\) are small compared to the ipsilateral parameters
\(b^i_v\), reflecting the low prevalence of contralateral lymph node
involvement for lateralized tumors.

@@

\caption{\label{tbl-midline-params}Mean sampled parameter estimates of
the midline model and the respective standard deviation.}

@@

\caption{}\label{T_0105a}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

\includegraphics[width=0.5\textwidth,height=\textheight]{manuscript_files/mediabag/figures/fig_model_midext_evo.pdf}

@@

The model introduced in the previous section defines a model of the
joint probability distribution over midline extension and ipsi- and
contralateral lymph node involvement,
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\).
How this quantity is computed is illustrated in
figure~\ref{fig-model-state-dist}, visually representing
equation~\ref{eq-bilateral-marginal-algebra}. To keep it manageable to
interpret, this example only considers the LNLs II, III, and IV ipsi-
and contralaterally, resulting in \(2^3 = 16\) distinct states per side
and thus \(2 \times 8 \times 8 = 128\) total states. LNLs I, V, and VII
and the associated spread parameters have been removed from the model
while the remaining parameters are set to the mean values in table xxx.
figure~\ref{fig-model-state-dist} shows the joint distribution as two
separate heatmaps for the two midline extension states for
early/advanced??? T-category tumors. The most likely state is a
lateralized tumor with involvement of ipsilateral level II and no
contralateral involvement, which has a probability of approximately
25\%. The second most probable state is a lateralized tumor with
ipsilateral involvement of levels II and III and no contralateral
involvement. The most likely state with contralateral involvement are
tumors with midline extension with involvement of contralateral level II
and ipsilateral level II and III.

@@

\includegraphics{manuscript_files/mediabag/figures/fig_model_state_dist.pdf}

@@

\caption{\label{fig-model-state-dist}Visual representation of
equation~\ref{eq-bilateral-marginal-algebra} for the case of midline
extention. The left and right matrix represent the evolution of the
possible hidden states for the ipsi- and contralateral neck
respectively. In particular, the right matrix describes the evolution of
the contralateral hidden states \emph{and} midline extension. In the
center, the time-prior is plotted as a diagonal matrix. The result of
this matrix computation is the joint distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\)}

@@

\includegraphics{manuscript_files/mediabag/figures/fig_model_prevalences_overall.pdf}

@@

\caption{\label{fig-model-prevalences-overall}Comparison of predicted
(histograms) vs observed (beta posteriors) prevalences. Shown for the
contralateral LNLs II (blue), III (orange), and IV (green). The top row
shows scenarios with early T-category tumors, the bottom row for late
T-category ones. The left column depicts scenarios where the primary
tumor is clearly lateralized, the right column scenarios of tumors
extending over the mid-sagittal line. This figure illustrates the
model's ability to describe the prevalence of involvement for different
combinations of the risk factors T-category and midline extension.}

@@

\includegraphics{manuscript_files/mediabag/figures/fig_model_prevalences_upstream.pdf}

@@

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

@@

\usepackage{bookmark}

@@

When treating head and neck squamous cell carcinomas (HNSCC) with
radiotherapy or surgery, not only the primary tumor and the clinically
detected lymph node metastases are targeted. Per current guidelines,
large volumes of the neck are part of the elective clinical target
volume (CTV-N)
\citep{gregoire_ct-based_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensity-modulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}.
This is to minimize the risk of regional recurrences due to untreated
microscopic disease which cannot be detected by current in-vivo imaging
modalities such as computed tomography (CT), magnetic resonance imaging
(MRI), or positron emission tomography (PET). However, minimizing the
risk of missing occult disease in the lymph drainage region must be
balanced against the toxicity related to unnecessary treatment of
healthy tissue.

The aforementioned guidelines for the definition of the CTV-N are based
on anatomically defined lymph node levels (LNL)
\citep{gregoire_delineation_2014} and the overall prevalence of lymph
node metastases in the LNLs. Based on this, they recommend extensive
irradiation of both sides of the neck for a majority of patients.
However, the overall prevalence of involvement in a given LNL does not
equal a patient's personal risk for occult disease in that LNL, which
depends on an individual patient's state of tumor progression.
Currently, a patient presenting with no clinically detectable nodal
disease (cN0) and a small, clearly lateralized T1 tumor, the same
recommended contralateral CTV-N as a patient with substantial nodal
involvement in the ipsilateral side and an advanced primary tumor that
crosses the mid-sagittal plane. Both patients receive extensive elective
irradiation of the contralateral LNLs II, III, and IVa
\citep{biau_selection_2019}.

To better quantify this personalized risk of occult disease we have
previously developed an intuitive, probabilistic hidden Markov model
\citep{ludwig_hidden_2021, ludwig_modelling_2023} which was originally
based on a conceptually similar Bayesian network model
\citep{pouymayou_bayesian_2019}. So far, these models were limited to
the ipsilateral side of the neck. The main contribution of this work is
the extension of the formalism to include risk predictions for the
contralateral side as well. The model may then guide the reduction of
the electively irradiated volumes in the contralateral neck for patients
whose individual risk for occult disease is low. In turn, this may
substantially reduce the toxicity from irradiation and the associated
decrease in quality of life.

@@

To be able to create models for lymphatic tumor progression that include
all relevant LNLs including the contralateral side, we have collected a
detailed dataset of 833 patients with newly diagnosed oropharyngeal
squamous cell carcinomas
\citep{ludwig_dataset_2022, ludwig_multi-centric_2023}. It reports the
lymph node involvement per LNL of every patient individually in tabular
form, in addition to other primary tumor and patient characteristics
such as T-category, subsite, lateralization of the primary tumor, and
HPV p16 status. Their patient records have been collected at four
different institutions and a brief overview over some of their patients'
characteristics are shown in table~\ref{tbl-data-overview}. The data
from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only
consist of patients who all received a neck dissection. The majority of
patients from the University Hospital Zürich (USZ) and the Hospital Vall
d'Hebron (HVH) was treated with definitive radiotherapy. Since sugical
treatment is more common for early T-category patients, the ISB and CLB
datasets contain a larger portion of early T-category patients compared
to the USZ and HVH dataset. For a subset of 83 patients in the CLB
dataset, the lateralization of the primary tumor was not reported.

@@

\caption{}\label{T_d2cbf}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

Pathological involvement is available only for the subset of surgical
patients and only for the dissected levels. For the remaining patients,
only clinical involvement is available. For the analysis performed in
this paper, all available diagnostic information has been combined into
a consensus decision representing the most likely state of involvement
for each patient and LNL, using literature values for sensitivity and
specificity of different modalities
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}. How
this consensus is formed is described in detail in
section~\ref{sec-consensus}. In brief, pathology after neck dissection
is taken as the ground truth if available, overwriting any contradicting
clinical involvement diagnoses. If a level was not dissected, the most
likely state is determined by PET-CT in most cases.

@@

The entire data, including additional patients with tumors in other
primary locations than the oropharynx, is publicly available: It may be
\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
it can be interactively explored too,
\href{https://github.com/rmnldwg/lydata}{from GitHub},
\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
\citet{ludwig_multi-centric_2023}. The Data-in-Brief publications
contain a detailed description of the datasets and the data format.
These publications do not include the most recent dataset addition from
HVH, which will be described in a future publication.

@@

These datasets allow us to investigate correlations between the
involvement of individual LNLs, or between risk factors and involvement.
In figure~\ref{fig-data-strat}, we have plotted the prevalence of each
contralateral LNL's involvement, stratified by the risk factors
T-category, number of ipsilaterally involved LNLs, and whether the tumor
extended over the mid-sagittal plane.

@@

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_data_strat.pdf}}

@@

The bottom left panel in figure~\ref{fig-data-strat} shows that patients
with a tumor crossing the mid-sagittal plane show contralateral
involvement vastly more often compared to patients with clearly
lateralized tumors. This makes intuitive sense, because the lymphatic
system in the head and neck region is typically symmetric and no major
lymph vessels cross the midline. Therefore, interstitial fluids from the
primary tumor -- which we assume to carry living malignant cells -- may
only reach the blind-ended lymphatic vessels in the contralateral neck
via short-ranged diffusion. Which in turn is only possible when the
primary tumor is close enough to the mid-sagittal line or crosses it.

The top left panel in figure~\ref{fig-data-strat} indicates that
T-category is correlated with contralateral involvement (as it is with
overal involvement). This is because T-category may be considered a
surrogate for the time between onset of disease and diagnosis. I.e., a
patient with a T4 tumor was -- on average -- diagnosed later than a
patient with a T1 tumor. Thus, the former did have more time to develop
metastases.

Similarly, ipsilateral involvement correlates with contralateral
metastasis (top right panel). The tumor of a patient with many
metastases in ipsilateral LNLs was probably able to spread for longer
(or faster) compared to a tumor in a patient with no nodal disease. The
degree of ipsilateral involvement, too, may therefore be considered a
surrogate for the duration of the disease. In addition, it has been
hypothesized that bulky nodal disease ipsilaterally may also redirect
lymph fluids to the contralateral side.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A tumor that extends over the mid-sagittal line should yield
  contralateral metastases with much higher probability.
\item
  More advanced T-category should lead to higher risk for nodal disease.
  One approach to achieve this via the expected time of diagnosis has
  already been developed in the form of a hidden Markov model
  \citep{ludwig_hidden_2021}.
\item
  The degree of ipsilateral involvement should give the model
  information on the time that may have passed between onset and
  diagnosis of the disease. This should come in addition to what can be
  inferred about the time from T-category alone.
\end{enumerate}

@@

\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{manuscript_files/mediabag/static/full-graph.pdf}

@@

Using equation~\ref{eq-bilateral-marginal} along with
equation~\ref{eq-lambda-matrix}, we can write the above distribution
algebraically as a product:

@@

\begin{equation}\phantomsection\label{eq-uni-bayes-law}{
P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid d, \boldsymbol{\hat{\theta}} \big)
= \frac{P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right) P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right) P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)}
{\sum_{i=0}^{2^V} \sum_{j=0}^{2^V} \mathcal{C}_{ij}}
}\end{equation}

with the normalization constants

\[
\mathcal{C}_{ij} = P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i \right) P \big( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j \big) P \big( \mathbf{X}^\text{i}=\boldsymbol{\xi}^\text{i}_i, \mathbf{X}^\text{c}=\boldsymbol{\xi}^\text{c}_j, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)
\]

@@

\subsection{Involvement Data Consensus}\label{sec-methods-consensus}

@@

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_burnin_history.pdf}}

@@

In table~\ref{tbl-midline-params}, we tabulate the mean and standard
deviation of the sampled parameters for the full midline model.
Considering the parameters of the bilateral model that describe the
ipsilateral spread (\(b^i_v\), \(t_{vr}\) and \(p_\text{adv.}\)), we
note that the bilateral model mostly reproduces the parameter values
reported in the earlier publication \citep{ludwig_modelling_2023} for
the unilateral model. The small differences may be mostly due to the
differences in the training dataset. Therefore, the analysis of the
model's capability to adequately describe ipsilateral lymph node
involvement is not repeated here and we focus instead on the analysis of
contralateral involvement. We can further note that the contralateral
spread parameters \(b^c_v\) are small compared to the ipsilateral
parameters \(b^i_v\), reflecting the low prevalence of contralateral
lymph node involvement for lateralized tumors.

@@

\caption{\label{tbl-midline-params}Mean sampled parameter estimates of
the midline model and the respective standard deviation. The parameters
set to fixed values are the maximum number of time steps \(t_{max}=10\)
and the time prior parameter for early T-category patients
\(p_{early}=0.3\).}

@@

\caption{}\label{T_7c6a5}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{manuscript_files/mediabag/figures/fig_model_midext_evo.pdf}

@@

The model introduced in the previous section defines a model of the
joint probability distribution over midline extension and ipsi- and
contralateral lymph node involvement,
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\).
How this quantity is computed is illustrated in
figure~\ref{fig-model-algebra-equation}, visually representing
equation~\ref{eq-bilateral-marginal-algebra}. To keep it manageable to
interpret, this example only considers the LNLs II, III, and IV ipsi-
and contralaterally, resulting in \(2^3 = 16\) distinct states per side
and thus \(2 \times 8 \times 8 = 128\) total states. LNLs I, V, and VII
and the associated spread parameters have been removed from the model
while the remaining parameters are set to the mean values in
table~\ref{tbl-midline-params}.

@@

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_algebra_equation.pdf}}

@@

\caption{\label{fig-model-algebra-equation}Visual representation of
equation~\ref{eq-bilateral-marginal-algebra} for the case of midline
extention. The left and right matrices represent the evolution of the
possible hidden states for the ipsi- and contralateral neck
respectively. In particular, the right matrices describe the evolution
of the contralateral hidden states once for the case of no midline
extension (top) and once for the case where the tumor does extend over
the mid-sagittal line (bottom). In the center, the time-prior for late
T-category is plotted as a diagonal matrix. The result of this matrix
computation is the joint distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)\),
visualized in figure~\ref{fig-model-state-dist}}

The matrix on the left side of figure~\ref{fig-model-algebra-equation}
shows the time evolution of the probability distribution over the
ipsilateral involvement state, starting from the healthy state
\([0,0,0]\) with probability one. The two matrices on the right side
show the time evolution of the probability distribution over the
contralateral involvement state, where we need to distinguish the two
cases with and without midline extension. At time step \(t=0\), the
contralateral neck starts from the healthy state \([0,0,0]\) without
midline extension. The matrix in the center shows the time prior for
late T-category. The result of the matrix multiplication is the joint
probability distribution over midline extension and ipsi- and
contralateral lymph node involvement,
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)\)
shown in figure figure~\ref{fig-model-state-dist}. The joint
distribution is shown as two separate heatmaps for the two midline
extension states. The most likely state is a lateralized tumor with
involvement of ipsilateral level II and no contralateral involvement,
which has a probability of approximately 25\%. The second most probable
state is a lateralized tumor with ipsilateral involvement of levels II
and III and no contralateral involvement. The most likely state with
contralateral involvement are tumors with midline extension with
involvement of contralateral level II and ipsilateral level II and III.

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_state_dist.pdf}}

\caption{\label{fig-model-state-dist}The full distribution
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\text{False} \right)\)
over the ipsi- and contralateral states, as well as midline extension as
computed by a model considering the LNLs II, III, and IV for a late
T-category tumor. Since this is a three-dimensional object with one
dimension being the binary random variable representing midline
extension \(\epsilon\), we show this distribution as separate
two-dimensional heatmaps. These two matrices are the result of
equation~\ref{eq-bilateral-marginal-algebra}, which we visualized as
well in figure~\ref{fig-model-algebra-equation}.}

@@

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_prevalences_overall.pdf}}

@@

\caption{\label{fig-model-prevalences-overall}Comparison of predicted
(histograms) vs observed (beta posteriors) prevalences, shown for the
contralateral LNLs II (blue), III (orange), and IV (green). The top row
shows scenarios with early T-category tumors, the bottom row for late
T-category ones. The left column depicts scenarios where the primary
tumor is clearly lateralized, the right column scenarios of tumors
extending over the mid-sagittal line. This figure illustrates the
model's ability to describe the prevalence of involvement for different
combinations of the risk factors T-category and midline extension.}

@@

\pandocbounded{\includegraphics[keepaspectratio]{manuscript_files/mediabag/figures/fig_model_prevalences_upstream.pdf}}

@@

\section{Consensus on most likely involvement}\label{sec-consensus}
